$p$-adic precision

### 2014-01-30 padicprec.tex
Differentials of several operations on matrices

%\documentclass{sig-alternate}
\documentclass{amsart}
\usepackage[utf8]{inputenc}

\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
\usepackage[algoruled,vlined,french,linesnumbered]{algorithm2e}

@@

\newtheorem{theo}{Theorem}[section]
\newtheorem{lem}[theo]{Lemma}
\newtheorem{prop}[theo]{Proposition}
\newtheorem{cor}[theo]{Corollary}
\newtheorem{quest}[theo]{Question}
\newtheorem{rem}[theo]{Remark}
\newtheorem{ex}[theo]{Example}
\theoremstyle{definition}
\newtheorem{deftn}[theo]{Definition}
\newtheorem{rmk}[theo]{Remark}

\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\R}{\mathbb R}
\renewcommand{\O}{\mathcal O}
\newcommand{\XX}{\mathbf X}

\newcommand{\val}{\text{\rm val}}
\renewcommand{\prec}{\text{\rm prec}}

@@

Two approaches are possible. The first one consists in doing the 
computation step-by-step: we first compute $(z,t) = f(x,y)$ and we 
compute then $f(z,t)$ which gives the result. Proceeding this way, we 
see that $z$ and $t$ are both known up to precision $O(p^{\min(n,m)})$ 
are consequently the same holds for the coordinates of the result.
The second approach consists in remarking that $f \circ f$ maps
$(x,y)$ to $(2x, 2y)$. Computing $f \circ f(x,y)$ using the latter
formula, it is clear that the result we get is known up to precision
$(O(p^n), O(p^m))$ --- and even $(O(p^{n+1}), O(p^{m+1}))$ if $p$
equals $2$.
As a conclusion, the second approach yields a more accurate result.
In other words, computing step-by-step creates ``artificial'' losses 
of precision.

@@

\section{Generalization to manifolds}

It may happen that natural $p$-adic ``objects'' do not lie in vector 
spaces. Actually, examples are numerous: points on the projective space 
or elliptic curves, subspaces of a fixed vector space (which lie in 
grassmannians), classes of isomorphism of certain curves (which lie in 
various moduli spaces), \emph{etc.} 
That is why we think that it is quite important to extend the formalism 
developped in Section \ref{sec:philosophy} to a more general setting. 
In this Section, we consider the quite general case of a strictly
differentiable manifolds locally modeled on (eventually infinite 
dimensional) ultrametric Banach spaces. This covers all the 
aforementioned examples.

@@

A little bit more interesting is the $n$-th power map $f$ from $K$ to 
itself. Its differential $f'(x)$ is obtained by differentiate the 
equality $y = x^n$: we obtain $dy = n x^{n-1} dx$.
Hence $f'(x)$ maps to ball $B_K(r)$ (where $r$ is some positive real 
number) to $B_K(|n|{\cdot} r)$. According to Lemma \ref{lem:main}, this 
means that the behaviour of the precision depends on the absolute value 
of $n$. By the ultrametric inegality, we always have $|n| \leq 1$ but 
this inegality might be strict. It happens for instance if the 
characteristic of $K$ is divisible by $p$, in which case $n$ vanishes in 
$K$. In that case $f'(x)$ also vanishes and Lemma \ref{lem:main} does 
not apply. 
Another situation (which is more interesting) is that of $p$-adic 
numbers: let us take $K = \Q_p$ and $n = p^k$ for some integer $k$. Then
$|n| = p^{-k}$ and Lemma \ref{lem:main} tells us basically that raising
a $p$-adic number to the $p^k$-th power increases the precision by $k$
extra digits (which is actually a well-known fact).

@@

We now focus on the computation of Gr\"obner basis. Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We 
can then consider the application mapping a family $(f_1, \ldots, f_s)$ 
of polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
application at $(f_1, \ldots, f_s)$ is continuous provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{Vaccon} for more precisions).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
Moreover following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
$$(d f_1, \ldots, d f_s) \cdot A = - (f_1, \ldots, f_s) \cdot dA + 
(d g_1, \ldots, d g_t).$$
Finally, using that we are working with \emph{reduced} Gr\"obner
basis, the above equality implies that $d f_i$ is the remainder in
the division of the $i$-th coefficient of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(f_1, \ldots, f_s)$.

@@

\todo{Determinant, characteristic polynomial, LU factorization, QR 
factorization, kernels, images}

@@

\todo{Intersection, sum}

@@

Recall (cf \S \ref{ssec:stepbystep}) that a SOMOS 4 sequence is a 
four-term inductive sequence defined by $u_{n+4} = \frac{u_{n+2} u_{n+4} 
+ u_{n+3}^3}{u_n}$. We recall also that SOMOS sequences exhibit the 
Laurent phenomenon: its means that the four initial terms $u_0, u_1, 
u_2, u_3$ are some indeterminates, that each $u_n$ is a Laurent 
polynomial with coefficients in $\Z$ in these indeterminates.

@@

= f \circ \cdots \circ f$ ($i$ times). Apparently $f$ is strictly 
differentiable on $\Q_p^\star \times \Q_p^3$ and its differential in the 

>>>>

%\documentclass{sig-alternate}
\documentclass{lms}
\usepackage[utf8]{inputenc}

\usepackage{amsmath,amssymb}
\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
\usepackage[algoruled,vlined,french,linesnumbered]{algorithm2e}

@@

\newtheorem{theo}{Theorem}[section]
\newtheorem{lem}[theo]{Lemma}
\newtheorem{prop}[theo]{Proposition}
\newtheorem{cor}[theo]{Corollary}
\newtheorem{quest}[theo]{Question}
\newtheorem{rem}[theo]{Remark}
\newtheorem{ex}[theo]{Example}
%\theoremstyle{definition}
\newtheorem{deftn}[theo]{Definition}
\newtheorem{rmk}[theo]{Remark}

\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\R}{\mathbb R}
\renewcommand{\O}{\mathcal O}
\newcommand{\XX}{\mathbf X}
\newcommand{\trans}{{}^{\text t}}

\newcommand{\val}{\text{\rm val}}
\newcommand{\tr}{\text{\rm Tr}}
\newcommand{\com}{\text{\rm Com}}
\renewcommand{\prec}{\text{\rm prec}}

@@

Two approaches are possible. The first one consists in doing the 
computation step-by-step: we first compute $(z,t) = f(x,y)$ and we 
compute then $f(z,t)$ which gives the result. Proceeding this way, we 
see that $z$ and $t$ are both known up to precision $O(p^{\min(n,m)})$ 
are consequently the same holds for the coordinates of the result.
The second approach consists in remarking that $f \circ f$ maps
$(x,y)$ to $(2x, 2y)$. Computing $f \circ f(x,y)$ using the latter
formula, it is clear that the result we get is known up to precision
$(O(p^n), O(p^m))$ --- and even $(O(p^{n+1}), O(p^{m+1}))$ if $p = 2$.
As a conclusion, the second approach yields a more accurate result.
In other words, computing step-by-step creates ``artificial'' losses 
of precision.

@@

\section{Generalization to manifolds}
\label{sec:manifold}

Many natural $p$-adic objects do not lie in vector spaces:
points in projective spaces or elliptic curves,
subspaces of a fixed vector space (which lie in Grassmannians),
classes of isomorphism of certain curves (which lie in various moduli spaces),
\emph{etc.}  In this section we extend the formalism 
developed in Section \ref{sec:philosophy} to a more general setting:
we consider the quite general case of strictly
differentiable manifolds locally modeled on ultrametric Banach spaces.
This covers all the 
aforementioned examples.

@@

A little bit more interesting is the $n$-th power map $f$ from $K$ to 
itself. Its differential $f'(x)$ is obtained by differentiate the 
equality $y = x^n$: we obtain $dy = n x^{n-1} dx$.
Hence $f'(x)$ maps to ball $B_K(r)$ (where $r$ is some positive real 
number) to $B_K(|n|{\cdot} r)$. According to Lemma \ref{lem:main}, this 
means that the behaviour of the precision depends on the absolute value 
of $n$. By the ultrametric inegality, we always have $|n| \leq 1$ but 
this inegality might be strict. It happens for instance if the 
characteristic of $K$ is divisible by $p$, in which case $n$ vanishes in 
$K$. In that case $f'(x)$ also vanishes and Lemma \ref{lem:main} does 
not apply. 
Another situation (which is more interesting) is that of $p$-adic 
numbers: let us take $K = \Q_p$ and $n = p^k$ for some integer $k$. Then
$|n| = p^{-k}$ and Lemma \ref{lem:main} reflects the well-known fact that
raising a $p$-adic number to the $p^k$-th power increases the precision 
by $k$ extra digits.

@@

We now focus on the computation of Gr\"obner basis. Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We can 
then consider the application mapping a family $(f_1, \ldots, f_s)$ of 
\emph{homogeneous}\footnote{This restriction will be convenient for us 
mainly because we are using the article \cite{Vaccon}, which deals only 
with homogeneous polynomials. It is however probably not essential.} 
polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from Theorem 1.1 of \cite{Vaccon} that this 
application at $(f_1, \ldots, f_s)$ is continuous provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is wearly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{Vaccon} for more precisions).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is a $(s \times t)$ matrix with coefficients in $K[\XX]$.
Moreover following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating Eq.~\eqref{eq:prodgrobner}, we get:
$$(d f_1, \ldots, d f_s) \cdot A = - (f_1, \ldots, f_s) \cdot dA + 
(d g_1, \ldots, d g_t).$$
Finally, using that we are working with \emph{reduced} Gr\"obner
basis, the above equality implies that $d f_i$ is the remainder in
the division of the $i$-th coefficient of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(f_1, \ldots, f_s)$.

@@

Differentiating ring operations (sum, products, inverse) over matrices 
is again straightforward; we just need to be careful that matrix 
algebras are (in general) noncommutative.

We now move to determinants, \emph{i.e.} we compute the differential
of the function $\det : M_n(K) \to K$. It is a standard computation.
To begin with, we consider an invertible matrix $M$ and write:
\begin{eqnarray*}
\det(M + dM) &=& \det(M) \cdot \det(I + M^{-1} \cdot dM)  \\
&=& \det(M) \cdot \big(1 + \tr(M^{-1} \cdot dM)\big) \\
&=& \det(M) + \tr(\com(M) \cdot dM)
\end{eqnarray*}
where $\com(M)$ denotes the comatrix of $M$. The differential of 
$\det$ at $M$ is then $dM \mapsto \tr(\com(M) \cdot dM)$. It turns
out that this formula is still valid when $M$ is not invertible.
The same computation extends readily to characteristic polynomials,
since they are defined as determinants. More precisely, let us 
consider the function $\chi : M_n(K) \to K_{\leq n}[X]$ (where
the latter ring denotes the finite dimensional vector space of 
polynomials over $K$ of degree at most $n$) taking a matrix $M$
to its characteristic polynomial $\det(X-M)$.
Then $\chi$ is differentiable at each point $M \in M_n(K)$ and its 
differential is given by $dM \mapsto \tr(\com(X{-}M) \cdot dM)$.

Other important tools related to matrices are usual factorizations as, 
for example, LU factorization. In what follows, let us agree to define a 
LU factorization of a square matrix $M \in M_n(K)$ as a writing $M = LU$ 
where $L$ is lower triangular and unipotent and $U$ is upper triangular. 
Such a writing exists and is unique provided that all principal minors 
of $M$ do not vanish. We can then consider the mapping $M \mapsto (L,U)$ 
defined over the Zariski-open set of matrices satisfying the above 
condition. In order to differentiate it, we differentiate the relation 
$M = LU$ and rewrite the result as follows:
$$L^{-1} dM \: U^{-1} = L^{-1} \cdot dL + dU \cdot U^{-1}.$$
We remark that in the right hand side of the above formula, the first
summand is lower triangular and all its diagonal entries vanish whereas 
the second summand is upper triangular. Hence in order to compute $dL$
and $dU$, one can proceed as follows: (1)~we compute the product $dX = 
L^{-1} dM \: U^{-1}$, (2)~we separate the lower and upper part of $dX$
obtaining this way $L^{-1} \cdot dL$ and $dU \cdot U^{-1}$ and (3)~we
recover $dL$ and $dU$ by multiplying the above matrices by $L$ on the 
left and $U$ on the right respectively.

A similar argument works for QR factorization.
For us, a QR factorization of a square matrix $M \in M_n(K)$ will be
a writing $M = QR$ where $R$ is unipotent upper triangular and $Q$ is
orthogonal in the sense that $\trans Q \cdot Q$ is diagonal. As before, such 
a writing exists and is unique on a Zariski-open subset of $M_n(K)$. The
mapping $f : M \mapsto (Q,R)$ is then well defined on this subset. We 
would like to emphasize at this point that the orthogonality condition
defines a sub-manifold of $M_n(K)$ which is a apparently \emph{not} a
vector space (it is defined by equations of degree $2$). The codomain
of $f$ is then also a manifold; this example then fits to the setting
of \S \ref{sec:manifold} but not to those of \S \ref{sec:philosophy}. 
The generalization made in \S \ref{sec:manifold} is then needed.
Anyway, we can differentiate $f$ by following the method we used for LU 
factorization: differentiating the relation $M = QR$, we obtain:
\begin{equation}
\label{eq:diffQR}
\trans Q \cdot dM \cdot R^{-1} = \trans Q \cdot dQ + \Delta \cdot dR 
\cdot R^{-1}
\end{equation}
where $\Delta = \trans Q \cdot Q$ is a diagonal matrix by definition.
Moreover by differentiating $\trans Q \cdot Q = \Delta$, we find that
$\trans Q \cdot dQ$ can be written as the sum of an antisymmetric 
matrix and a diagonal one. Since moreove $dR \cdot R^{-1}$ is upper
triangular with all diagonal entries equal to $0$, we see that 
Eq.~\eqref{eq:diffQR} is enough to compute $dQ$ and $dR$ from $Q$, $R$ 
and $dM$.

@@

\todo{Kernels, images, intersection, sum}

@@

Recall (\emph{cf} \S \ref{ssec:stepbystep}) that a SOMOS 4 sequence is a 
four-term inductive sequence defined by $u_{n+4} = \frac{u_{n+2} u_{n+4} 
+ u_{n+3}^3}{u_n}$. We recall also that SOMOS sequences exhibit the 
Laurent phenomenon: its means that the four initial terms $u_0, u_1, 
u_2, u_3$ are some indeterminates, that each $u_n$ is a Laurent 
polynomial with coefficients in $\Z$ in these indeterminates.

@@

= f \circ \cdots \circ f$ ($i$ times). Clearly, $f$ is strictly 
differentiable on $\Q_p^\times \times \Q_p^3$ and its differential in the 
